Paleoenvironmental models for Australia and the impact of aridification on blindsnake diversification

### 2023-07-06 manuscript.Rmd
change citations

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo of their diversification. In particular, one major aim of evolutionary biology is to understand the role that environmental factors have played in shaping rates of species diversification [@simpsonTempo1944]. A variety of environmental correlates have been associated with diversification processes across clades, including elevation [@esquerreHow2019;@boschmanMountain2022], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017], temperature [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014]. However, despite advancements in our understanding of the evolutionary relationships and divergence times of many clades, our knowledge of the impact of paleoenvironmental change on these clades remains limited, particularly in regard to the regional scale at which many clades radiate [@eyresClimatic2021]. As such, it remains a key challenge to combine regional paleoenvironmental histories, with molecular phylogenies, to understand the drivers of diversification over deep-time scales.

@@

The Australian continent has one of the most diverse reptile faunas on the planet and this may have been driven by its rapid environmental change in the Neogene period. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annum) with the development of pockets of aridity [@martinCenozoic2006;@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, as aridification progressed, arid-adapted vegetation expanded while mesic biomes became increasingly fragmented [@crispHow2013;@martinCenozoic2006]. This shift in the environment also led to a radiation of some reptile clades, such as elapid snakes and skinks, which diversified into morphologically and ecologically distinct lineages [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Conversely, other reptile clades, such as blindsnakes and geckos, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. Furthermore, shifts in diversification rates due to changes in Australian temperature and aridity have been reported in some squamate reptile clades including skinks [@raboskyExceptional2007] and geckos [@fujitaDiversification2010;@oliverContrasting2014;@oliverIndependent2014;@pepperGenetic2011;@pepperPalaeoclimate2011] as well in frogs [@catulloAridification2014;@vidal-garciaConvergent2015], insects [@owenHow2017;@toussaintUnveiling2015], and plants [@crispHow2013;@kadereitMolecular2010;@millerComparative2013]. Yet, the role of aridification and temperature changes on diversification is highly variable, with some arid-adapted vertebrates radiating following aridification [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] while other mesic-adapted groups have had suppressed diversification or been driven extinct [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017]. Whether environmental change in Australia has had generally positive or negative influence on diversification rates in squamate reptiles is still unknown.

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species **47** described species [@uetzQuarter2021] (Tiatragul et al., accepted). All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from Tiatragul et al. (accepted) indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin.
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower radiation.

@@

We used the most updated phylogeny for *Anilios* and closely related outgroups from Tiatragul et al., submitted MS. The species tree was built with the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] using maximum likelihood gene trees inferred in IQTREE from 4930 loci (322 Anchored Hybrid Enrichment, 4574 ultraconserved elements, and 34 legacy genes) for 35/47 recognised *Anilios* species. Tiatragul et al. (accepted) time-calibrated the phylogeny using 27 outgroup squamate fossils as primary calibrations using MCMCTree from the PAML package [@yangPAML2007]. Phylogenetic data in Tiatragul et al. (submitted) suggest the existence of at least three genetically distinct undescribed species, so we followed their designation and adjusted our taxon sampling fraction to 38/50 species. 
The genetically distinct OTUs that are not fully described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and two '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) (Tiatragul et al., accepted). 
For ancestral biome estimation analyses, we included three outgroup blindsnakes species from Southeast Asia, Indonesia, and Melanesia, where the ancestor of *Anilios* has been hypothesised to have originated [@vidalBlindsnake2010; Tiatragul et al., accepted]. The three species include *Ramphotyphlops multilineatus* (Western New Guinea, Kai Island, and Salawati), *Sundatyphlops polygrammicus* (Moyo Island, Indonesia), and *Acutotyphlops subocularis* (Bismarck Archipelago). We were not able to include a *Ramphotyphlops* from continental Southeast Asia in the time-calibrated phylogeny because insufficient data was available for divergence time estimation.

@@

Tiatragul et al., (accepted) proposed that the earliest divergence of *Anilios* occurred around 25 Ma (28 – 22 Ma; 95% HPD), which is slightly older than 21.9 Ma (30.2 – 15.1 Ma; 95% Bayesian credibility interval) proposed by previous studies [@marinTracing2013;@vidalBlindsnake2010]. Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the tropical grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of widespread aridification in the late Miocene [@byrneBirth2008].

@@

The observed decrease in diversification rate with increasing aridification may be linked to decreasing speciation rates or increasing extinction rates. One hypothesis is that mesic-adapted ancestral blindsnakes may have been widespread in the past, but populations later became extinct due to extreme aridity in the last 5 Ma [@fujiokaAustralian2009;@byrneDecline2011].
An extensive population genetics study of blindsnake by @marinHidden2013 proposed that during widespread aridification, mesic-adapted lineages went extinct or were confined to mesic refugia, such as the Pilbara and Kimberley, where they became pre-adapted to seasonal aridity, which enable subsequent invasion into the truly arid environments.
Additionally, long naked branches in our current phylogeny, particularly in Group 1, suggests past extinction events or intraspecific diversity not yet recognised.
Our best-fitting models of diversification support constant extinction rates with decreasing rates of speciation. 
However, the lack of *Anilios* fossils makes extinction rates difficult to estimate [@loucaWhy2021;@raboskyExtinction2010]. An alternative explanation for 
the negative relationship between aridification and diversification is decreasing speciation rates through time. This could be driven by the indirect effect of aridification on population size [@goldieEvolutionary2010]. More arid sites have lower primary productivity and may support smaller population sizes. Under the integrated evolutionary speed hypothesis [@gillmanPatterns2013;@rohdeLatitudinal1992], smaller population sizes are predicted to have slower rates of molecular evolution and subsequently lower rates of speciation. Lower rates of molecular evolution have been found in Australian arid zone plants compared to mesic zone plants [@goldieEvolutionary2010]. 
Yet, here we found no difference between arid zone and mesic zone speciation rates in blindsnakes which is inconsistent with this hypothesis. 
Finally, another potential explanation for the negative correlation between diversification and aridity is the limited sampling of Australian blindsnakes, particularly in major sand deserts [@pepperLife2021] or even in areas of high cryptic diversity such as the Kimberley [@ellisTyphlopid2019]. The current phylogeny is missing 12 described species (Tiatragul et al., submitted MS), such as the arid-adapted _A. fossor_ and _A. minimus_, and the true diversity has been estimated to exceed 56 species [@marinHidden2013;@marinTracing2013;@ellisTyphlopid2019]. It is possible that unrecognised species may be hidden under nominal species and that these missing lineages have diversified more recently when aridity is high. If all suspected cryptic OTUs were treated as recognised species, we could see a different relationship with aridification.

>>>>

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo of their diversification. In particular, one major aim of evolutionary biology is to understand the role that environmental factors have played in shaping rates of species diversification [@simpsonTempo1944]. A variety of environmental correlates have been associated with diversification processes across clades, including elevation [@esquerreHow2019;@boschmanMountain2022], CO<sub>2</sub> concentration [@kergoatOpposite2018], microhabitats [@bars-closelDiversification2017], temperature [@ricklefsGlobal2006;@mayhewBiodiversity2012;@condamineMacroevolutionary2013;@condamineClimate2019;@rollandContribution2019;@sunRecent2020], and aridification [@raboskyExceptional2007;@pepperGenetic2011;@deepakAridification2018;@catulloAridification2014]. However, despite advancements in our understanding of the evolutionary relationships and divergence times of many clades, our knowledge of the impact of paleoenvironmental change on these clades remains limited, particularly in regard to the regional scale at which many clades radiate [@eyresClimatic2021]. As such, it remains a key challenge to combine regional paleoenvironmental histories with molecular phylogenies to understand the drivers of diversification over deep-time scales.

@@

The Australian continent has one of the most diverse reptile faunas on the planet and this may have been driven by its rapid environmental change in the Neogene period. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annum) with the development of pockets of aridity [@martinCenozoic2006;@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, as aridification progressed, arid-adapted vegetation expanded while mesic biomes became increasingly fragmented [@crispHow2013;@martinCenozoic2006]. This shift in the environment also led to a radiation of some reptile clades, such as elapid snakes and skinks, which diversified into morphologically and ecologically distinct lineages [@leeDiversification2016;@sandersMolecular2008;@raboskyExceptional2007]. Conversely, other reptile clades, such as blindsnakes and geckos, remain morphologically conserved and were thought to be species poor until recent molecular data suggested otherwise [@moritzCryptic2018;@zozayaAre2019;@marinHidden2013]. Furthermore, shifts in diversification rates due to changes in Australian temperature and aridity have been reported in some squamate reptile clades including skinks [@raboskyExceptional2007] and geckos [@fujitaDiversification2010;@oliverContrasting2014;@oliverIndependent2014;@pepperGenetic2011;@pepperPalaeoclimate2011] as well in frogs [@catulloAridification2014;@vidal-garciaConvergent2015], insects [@owenHow2017;@toussaintUnveiling2015], and plants [@crispHow2013;@kadereitMolecular2010;@millerComparative2013]. Yet, the role of aridification and temperature changes on diversification is highly variable, with some arid-adapted vertebrates radiating following aridification [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] while other mesic-adapted groups have had suppressed diversification or been driven to extinction [@bryantLines2016;@cardilloPhylogeny2017;@owenHow2017]. Whether environmental change in Australia has had generally positive or negative influence on diversification rates in squamate reptiles is still unknown.

Today, the arid zone covers roughly 70% of the Australian landscape and provides habitat to 43% of all of Australia’s terrestrial reptiles [@byrneBirth2008].Of these, roughly a quarter of the snake species diversity in Australia is represented by blindsnakes of the genus *Anilios*, which currently comprises 47 described species **47** described species [@uetzQuarter2021;@tiatragulAustralia2023]. All but three species of currenty described _Anilios_ are endemic to Australia, _A. torresianus_ and _A. leucoproctus_ also occur in southern parts of New Guinea, while _A. eryncinus_ has only been recorded in New Guinea and the Maluku Islands. 
Blindsnakes are morphologically conserved and previous phylogenetic and morphological studies have revealed that the true species diversity is underestimated [@marinHidden2013;@pyronSystematics2014;@hedgesTaxonomic2014;@ellisTyphlopid2019]. The diversification of _Anilios_ in Australia is reported to have originated from a single dispersal event by a mesic-adapted ancestor from Southeast Asia or Indonesia between 39 – 19 Ma [@vidalBlindsnake2010;@marinTracing2013]. 
However, phylogenetic analyses from @tiatragulAustralia2023 indicated that the closest relative to all *Anilios* is *Ramphotyphlops multilineatus*, a species that occur in Western New Guinea, Kai Islands, and Salawati, suggesting the Australian radiation may have a Melanesian origin.
Despite representing a substantial portion of snake diversity, we know little about the macroevolutionary dynamics of blindsnakes. The monophyly of the genus makes _Anilios_ well suited to test hypotheses about how lineages diversified with dramatic environmental changes during the Cenozoic because they are geographically restricted and can be found in every biome on mainland Australia.

@@

In this study, we aimed to infer the historical biogeography of the Australian blindsnake radiation and investigate the impact of paleoenvironmental changes, specifically changes in temperature and aridity associated with Neogene aridification, on diversification rates within this group. To do this, we fitted several birth-death models with reconstructed paleotemperature and aridity, and estimated diversification rates under these models. Additionally, we compared diversification rates between arid-adapted lineages versus mesic-adapted lineages to investigate if lineages that inhabit the arid biome have undergone faster or slower diversification.

@@

We used the most updated phylogeny for *Anilios* and closely related outgroups from Tiatragul et al., submitted MS. The species tree was built with the shortcut coalescent method in ASTRAL III [@zhangASTRALIIIPolynomialTime2018] using maximum likelihood gene trees inferred in IQTREE from 4930 loci (322 Anchored Hybrid Enrichment, 4574 ultraconserved elements, and 34 legacy genes) for 35/47 recognised *Anilios* species. @tiatragulAustralia2023 time-calibrated the phylogeny using 27 outgroup squamate fossils as primary calibrations using MCMCTree from the PAML package [@yangPAML2007]. Phylogenetic data in @tiatragulAustralia2023 suggest the existence of at least three genetically distinct undescribed species, so we followed their designation and adjusted our taxon sampling fraction to 38/50 species. 
The genetically distinct OTUs that are not formally described species include three '_Anilios grypus_' (two in Western Australia and one in Queensland) and two '_Anilios ligatus_' (one from the eastern part of the range and one from the northern part) [@tiatragulAustralia2023]. 
For ancestral biome estimation analyses, we included three outgroup blindsnakes species from Southeast Asia, Indonesia, and Melanesia, where the ancestor of *Anilios* has been hypothesised to have originated [@vidalBlindsnake2010; Tiatragul et al., accepted]. The three species include *Ramphotyphlops multilineatus* (Western New Guinea, Kai Island, and Salawati), *Sundatyphlops polygrammicus* (Moyo Island, Indonesia), and *Acutotyphlops subocularis* (Bismarck Archipelago). We were not able to include a *Ramphotyphlops* from continental Southeast Asia in the time-calibrated phylogeny because insufficient data was available for divergence time estimation.

@@

@tiatragulAustralia2023 proposed that the earliest divergence of *Anilios* occurred around 25 Ma (28 – 22 Ma; 95% HPD), which is slightly older than 21.9 Ma (30.2 – 15.1 Ma; 95% Bayesian credibility interval) proposed by previous studies [@marinTracing2013;@vidalBlindsnake2010]. Our BioGeoBEARS analyses suggest that _Anilios_ gradually diversified in the tropical grassland (Group 1) and arid biome (Group 2) during a time when Australia was relatively wet and humid [@byrneBirth2008]. This indicates that many species found in arid biomes, particularly in Group 2, diversified in the arid biome prior to the onset of widespread aridification in the late Miocene [@byrneBirth2008].

@@

The observed decrease in diversification rate with increasing aridification may be linked to decreasing speciation rates or increasing extinction rates. One hypothesis is that mesic-adapted ancestral blindsnakes may have been widespread in the past, but populations later became extinct due to extreme aridity in the last 5 Ma [@fujiokaAustralian2009;@byrneDecline2011].
An extensive population genetics study of blindsnake by @marinHidden2013 proposed that during widespread aridification, mesic-adapted lineages went extinct or were confined to mesic refugia, such as the Pilbara and Kimberley, where they became pre-adapted to seasonal aridity, which enable subsequent invasion into the truly arid environments.
Additionally, long naked branches in our current phylogeny, particularly in Group 1, suggests past extinction events or intraspecific diversity not yet recognised.
Our best-fitting models of diversification support constant extinction rates with decreasing rates of speciation. 
However, the lack of *Anilios* fossils makes extinction rates difficult to estimate [@loucaWhy2021;@raboskyExtinction2010]. An alternative explanation for 
the negative relationship between aridification and diversification is decreasing speciation rates through time. This could be driven by the indirect effect of aridification on population size [@goldieEvolutionary2010]. More arid sites have lower primary productivity and may support smaller population sizes. Under the integrated evolutionary speed hypothesis [@gillmanPatterns2013;@rohdeLatitudinal1992], smaller population sizes are predicted to have slower rates of molecular evolution and subsequently lower rates of speciation. Lower rates of molecular evolution have been found in Australian arid zone plants compared to mesic zone plants [@goldieEvolutionary2010]. 
Yet, here we found no difference between arid zone and mesic zone speciation rates in blindsnakes which is inconsistent with this hypothesis. 
Finally, another potential explanation for the negative correlation between diversification and aridity is the limited sampling of Australian blindsnakes, particularly in major sand deserts [@pepperLife2021] or even in areas of high cryptic diversity such as the Kimberley [@ellisTyphlopid2019]. The current phylogeny is missing 12 described species [@tiatragulAustralia2023], such as the arid-adapted _A. fossor_ and _A. minimus_, and the true diversity has been estimated to exceed 56 species [@marinHidden2013;@marinTracing2013;@ellisTyphlopid2019]. It is possible that unrecognised species may be hidden under nominal species and that these missing lineages have diversified more recently when aridity is high. If all suspected cryptic OTUs were treated as recognised species, we could see a different relationship with aridification.
